### 2016-07-03 case-studies.tex
edits

In this section, we present two case studies that demonstrate the
applicability of our reasoning approach in varied settings, and the
consequent practical utility.

\paragraph{A video view counter} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
tools we developed to analyze Ruby can be found at the following
anonymized url: \url{...}}. The application stores the video count in
a Postgres database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default {\sc rc}
isolation of Postgres may lead to anomalous executions as confirmed by
our experiments; with 16 (resp.  32) concurrent writers performing
single increments, we observed a total of 7 (17) violations of
monotonicity. We subsequently instantiated our reasoning framework for
an {\sc sc} store with Postgres isolation levels (which are stronger
than ANSI SQL's and MySQL's), and spent 5 man hours determining
appropriate isolation levels needed to enforce the
invariant\footnote{The representative \txnimp programs we used to
formally reason about scenarios described in this section can be found
in the supplementary.} In particular, we were able to formally prove
that to preserve the invariant, writers need to execute at
\iso{Repeatable Read} isolation, while readers can continue at {\sc
rc}. Repeating the experiments with this configuration led to no
violations, as expected. While executing all transactions at {\sc ser}
also prevents violations, it led to an additional 47\% increase in
latency on an average (16 concurrent writers; 5 rounds). We repeated
the exercise for MySQL, and found that its {\sc rr} is not sufficient
to enforce monotonicity; {\sc ser} is needed.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency, we
reimplemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency with {\sc cc} allowed us to obtain a proof, and
recover the monotonicity invariant. We in fact obtained a proof after
relaxing the serializability requirement for writes, by changing the
definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return the
value of the largest write-to-$X$ instead of the latest write-to-$X$.
We implemented this version of the counter RDT, and obtained a 52\%
improvement in latency compared to the previous
implementation\footnote{The tradeoff is that the view count is now no
longer accurate, although it grows monotonically.}.

>>>>

In this section, we present two case studies that demonstrate the
applicability of our reasoning approach in varied settings, and its
consequent practical utility.

\paragraph{A video view counter} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
  tools we developed to analyze Ruby can be found at the following
  anonymized url: \url{...}}. The application stores the video count
in a Postgres database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default {\sc rc}
isolation of Postgres may lead to anomalous executions as confirmed by
our experiments; with 16 (resp.  32) concurrent writers performing
single increments, we observed a total of 7 (17) violations of
monotonicity. We subsequently instantiated our reasoning framework for
an {\sc sc} store with Postgres isolation levels (which are stronger
than ANSI SQL's and MySQL's), and spent 5 man hours determining
appropriate isolation levels needed to enforce the
invariant\footnote{The representative \txnimp programs we used to
  formally reason about scenarios described in this section can be
  found in the supplementary material.} In particular, we were able to
formally prove that to preserve the invariant, writers need to execute
at a \iso{Repeatable Read} isolation level, while readers can continue
to operate using \iso{Read Committed} isolation.  While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 47\% increase in latency on an average (16 concurrent
writers; 5 rounds). We repeated the exercise for MySQL, and found that
using {\sc rr} was not sufficient to enforce monotonicity, but that
{\sc ser} was needed, due to differences in how {\sc rr} is
implemented between the two systems.

% To understand whether a straightforward implementation of the view
% counter preserves the invariant, we considered a \txnimp program
% that is representative of the counter application. In particular, we
% considered four concurrent transactions that operate on a shared
% counter: a reader transaction that reads the counter twice obtaining
% the values $k_1$ and $k_2$, respectively, and three writer
% transactions, each incrementing the counter once. If the reader
% witnesses a monotonically increasing view count, then $k_2\ge k_1$.
% We then instantiated our reasoning framework for an {\sc sc} store
% with {\sc rc} as the default isolation level (Postgres is one such
% store), with the aim of verifying the program under this
% instantiation.  Our attempts were however unsuccessful owing to our
% inability to prove that a writer transaction always commits a value
% no smaller than the current value of the counter. This points to the
% existence of anamolous executions where  To understand the
% prevalence of such anomalies, we implemented a view counter
% application in Ruby-on-Rails using standard templates and
% techniques.   
To understand the effect of changing the store consistency level, we
reimplemented the counter as a \emph{replicated data type}
(RDT)~\cite{crdt,burckhardt14} on top of Cassandra
\emph{bolted-on}~\cite{bailisbolton} with {\sc cc} and {\sc sc}
consistency levels, and ANSI SQL isolation levels. Keeping the
consistency level at {\sc ec}, we then repeated our experiments with
committed reads and serializable increments - a combination which was
proved to preserve monotonicity on an {\sc sc} store, but nonetheless
observed monotonicity violations. On the other hand, instantiating the
store consistency level with {\sc cc} allowed us to obtain a proof,
and recover the monotonicity invariant. We in fact obtained a proof
after relaxing the serializability requirement for writes, by changing
the definition of the $\interp{\E}(X)$ (\S\ref{sec:opsem}) to return
the value of the largest write-to-$X$ instead of the latest
write-to-$X$.  We implemented this version of the counter RDT, and
obtained a 52\% improvement in latency compared to the previous
implementation\footnote{The tradeoff is that the view count is now no
  longer accurate, although it grows monotonically, thus preserving
  the application invariant.}.

### 2016-07-03 related-work.tex
edits

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specified several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anomaly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace. Cerone \emph{et al.}~\cite{gotsmanconcur15} present a framework
to specify isolation levels with atomic visibility, but their
specifications are also for complete traces; Moreover, their
formulation implicitly assumes homogeneity, i.e., they require all
transactions to execute using either {\sc si} or {\sc ser}, which is
often not the case in practice. While all the aforementioned
specification frameworks use the vocabulary introduced
in~\cite{burckhardt14} to specify replicated data types and different
forms of weak consistency, in contrast to this work, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

\paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
\emph{et al.} propose a theory to characterize non-serializable executions
that arise under {\sc si}. Fekete~\cite{fekete2005} also proposes an
algorithm that allocates either {\sc si} or {\sc ser} isolation levels
to transactions while guaranteeing the serializability of the
execution. In~\cite{gotsmanpodc16}, Cerone \emph{et al.} improve on Adya's
{\sc si} specification and use it to derive a static analysis that
determines the safety of substituting {\sc si} with a weaker variant
called \iso{Parallel Snapshot Isolation}~\cite{psi}. All of them focus
on establishing the equivalence of executions between a pair of
isolation levels, without taking application invariants into account.
Bernstein \emph{et al.}~\cite{bern2000} propose informal semantic conditions
to ensure the satisfaction of application invariants under weaker
isolation levels.  However, no formal proof system has been proposed.
Furthermore, all the aforementioned approaches are tailor-made for a
finite set of well-understood isolation levels (rooted
in~\cite{berenson}) under a pre-defined consistency model (usually
{\sc sc}).

\paragraph{Reasoning under weak consistency} There have been several
recent proposals on reasoning techniques for programs executing under
weak consistency~\cite{bailisvldb, alvarocalm,
  gotsmanpopl16,redblueatc, redblueosdi, ecinec}. All of them assume a
system model that offers a choice between a \emph{coordination-free}
weak consistency level (\emph{e.g.}, {\sc ec}~\cite{redblueosdi,
  redblueatc, ecinec, alvarocalm, bailisvldb} or causal consistency
({\sc cc})~\cite{gotsmanpopl16}). In all these efforts, the key
technical issue involves proving that atomic and fully isolated
operations preserve application invariants when executed on these
consistency levels.  In contrast, we admit weakly isolated
transactions (sets of atomic operations), and our system model accepts
\emph{specifications} of consistency and isolation levels drawn from
an expressive logic.  In recent work~\cite{gotsmanpopl16},
\iso{Parallel Snapshot Isolation} is adapted to the aforementioned
setting by interpreting it as a consistency level that serializes
writes to objects; a dedicated proof rule is developed to help prove
prove program invariants hold under this model. By parameterizing our
proof system over a gamut of weak isolation specifications, we avoid
the need to define a separate proof rule for each new isolation level.

\paragraph{Reasoning under relaxed memory} Considering the
correspondences between axiomatic models of relaxed
memory~\cite{battycpp} and weakly consistent
replication~\cite{burckhardt14}, it is reasonable to expect similar
reasoning approaches to work in both the cases. Ridge~\cite{rgtso}
generalizes rely-guarantee reasoning to the x86-TSO memory model.
Likewise, Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent
separation logic (CSL)~\cite{csl} to the C11 relaxed memory model.
Ferreira \emph{et al.}~\cite{ferreira10} propose a parameterized
operational semantics for relaxed memory models, but the
parameterization is over a relation between relaxed memory programs
and related {\sc sc} programs. They also prove the soundness of CSL
for the weakest memory model expressible in their semantics, but
neither a parameterized (in the aforementioned sense) CSL, nor a CSL
specialized for any relaxed memory model is proposed.  ~\cite{DLZ+13}
presents a \emph{buffered memory model} for Java that defines an
axiomatic definition for the JMM in terms of memory reorderings, and
an operational instantiation consistent with the TSO memory model
found on x86 processors, and demonstrates the equivalence of the two
definitions.  However, there is no attempt to parameterize the model
over different reordering or consistency definitions, which is a central
focus of this paper.

>>>>

\paragraph{Specifying weak isolation.}
Adya~\cite{adyaphd} specifies several weak isolation levels in terms
of \emph{dependency graphs} between transactions, by defining the
kinds of dependency cycles that are forbidden in each case. The
operational nature of Adya's specifications make them suitable for
runtime monitoring and anamoly
detection~\cite{kemmevldb,feketesigmod08,pssi2011}, whereas the
declarative nature of our specifications make them suitable for formal
reasoning about program behaviour. Sivaramakrishnan \emph{et
  al.}~\cite{pldi15} specify isolation levels declaratively as trace
well-formedness conditions, but their specifications implicitly assume
a complete trace with only committed transactions, and is thus not
suitable to reason operationally about a program as it builds the
trace; their framework also requires application invariants be
expressed in terms of the visibility relations used to define
isolation and consistency properties.  Cerone \emph{et
  al.}~\cite{gotsmanconcur15} present a framework to specify isolation
levels with atomic visibility, but their specifications are also for
complete traces; moreover, their formulation implicitly assumes
homogenity, \emph{i.e.}, they require all transactions to execute
using either {\sc si} or {\sc ser}, which is often not the case in
practice. While all the aforementioned specification frameworks use
the vocabulary introduced in~\cite{burckhardt14} to specify replicated
data types and different forms of weak consistency, none of them come
with an associated reasoning framework that can use such
specifications productively towards verifying programs under weak
isolation.

\paragraph{Reasoning under weak isolation} In~\cite{feketessi}, Fekete
\emph{et al.} propose a theory to characterize non-serializable
executions that arise under {\sc si}. Fekete~\cite{fekete2005} also
proposes an algorithm that allocates either {\sc si} or {\sc ser}
isolation levels to transactions while guaranteeing the
serializability of the execution. In~\cite{gotsmanpodc16}, Cerone
\emph{et al.} improve on Adya's {\sc si} specification and use it to
derive a static analysis that determines the safety of substituting
{\sc si} with a weaker variant called \iso{Parallel Snapshot
  Isolation}~\cite{psi}.  These efforts focus on establishing the
equivalence of executions between a pair of isolation levels, without
taking application invariants into account.  Bernstein \emph{et
  al.}~\cite{bern2000} propose informal semantic conditions to ensure
the satisfaction of application invariants under weaker isolation
levels.  However, no formal proof system is given.  Furthermore, these
techniques are tailor-made for a finite set of well-understood
isolation levels (rooted in~\cite{berenson}) under a pre-defined
store consistency model (usually {\sc sc}).

\paragraph{Reasoning under weak consistency} There have been several
recent proposals on reasoning techniques for programs executing under
weak consistency~\cite{bailisvldb, alvarocalm,
  gotsmanpopl16,redblueatc, redblueosdi, ecinec}. All of them assume a
system model that offers a choice between a \emph{coordination-free}
weak consistency level (\emph{e.g.}, {\sc ec}~\cite{redblueosdi,
  redblueatc, ecinec, alvarocalm, bailisvldb}) or causal
consistency~\cite{gotsmanpopl16}). The key technical issue in these
efforts involves proving that atomic and fully isolated operations
preserve application invariants when executed on these consistency
levels.  In contrast, we admit weakly isolated transactions, and our
system model accepts \emph{specifications} of consistency and
isolation levels drawn from an expressive logic.  In recent
work~\cite{gotsmanpopl16}, \iso{Parallel Snapshot Isolation} is
adapted to the aforementioned setting by interpreting it as a
consistency level that serializes writes to objects; a dedicated proof
rule is developed to help prove prove program invariants hold under
this model. By parameterizing our proof system over a gamut of weak
isolation specifications, we avoid the need to define a separate proof
rule for each new isolation level we may encounter.

\paragraph{Reasoning under relaxed memory} The reasoning mechanisms
used to describe and prove properties about weakly-isolated
transactions bear resemblance to those used to formalize relaxed
memory behaviour~\cite{battycpp}.  Ridge~\cite{rgtso} generalizes
rely-guarantee reasoning to the x86-TSO memory model.  Likewise,
Vafeiadis \emph{et al.}~\cite{rsl13} generalize concurrent separation
logic (CSL)~\cite{csl} to the C11 relaxed memory model.  Ferreira
\emph{et al.}~\cite{ferreira10} propose a parameterized operational
semantics for relaxed memory models, but the parameterization is over
a relation between relaxed memory programs and related {\sc sc}
programs. They also prove the soundness of CSL for the weakest memory
model expressible in their semantics, but neither a parameterized (in
the aforementioned sense) CSL, nor a CSL specialized for any relaxed
memory model is proposed.  ~\cite{DLZ+13} presents a \emph{buffered
  memory model} for Java that defines an axiomatic definition for the
JMM in terms of memory reorderings, and an operational instantiation
consistent with the TSO memory model found on x86 processors, and
demonstrates the equivalence of the two definitions.  However, there
is no attempt to parameterize the model over different reordering or
consistency definitions, which is a central focus of this paper.
